White wine analysis by Ilaria Tavecchia
========================================================
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.
library(knitr)
library(ggplot2)
library(GGally)
library(dplyr)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
pf <- read.csv('wineQualityWhites.csv')
pf$X <- NULL
#names(pf)

```

#Information about the data
This dataset contains information about white wine. This tidy dataset contains 
4,898 white wine with 11 variables on quantifying the chemical properties of 
each wine. At least 3 wine experts rated the quality of each wine, providing a 
rating between 0 (very bad) and 10 (excellent).

# Univariate Plots Section

First of all, let's have a look at the data. All our columns are numeric, as 
well as the quality column that contains the wine evaluation. We have 
added a new column called quality_cat, that summarises the quality in 3 values: 
Low for quality below 4, medium for quality between 5 and 8 and high for 
everything above 8. Further in our analysis we will just focus on Low and High, 
considering that they contain roughly the same amount of values. Some descriptive
statistics are shown as well below. 

```{r Univariate_Plots}
#create a variable as an ordered factor
pf$quality_cat <-ifelse(pf$quality<=4, "Low", 
                        ifelse(pf$quality>=8, "High", "Medium"))

pf$quality_cat <-ordered(pf$quality_cat, levels = c("Low", "Medium", "High"))

summary(pf)


```


First of all, let's have a look at the distribution for all our variables.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <- qplot(data =pf, x = volatile.acidity) + 
  xlab("volatile acidity (acetic acid - g / dm^3)")
p2 <- qplot(data =pf, x = citric.acid) + xlab("citric acid (g / dm^3)")
p3 <- qplot(data =pf, x = residual.sugar) + xlab("residual sugar (g / dm^3)")
p4 <- qplot(data =pf, x = chlorides) + xlab("chlorides (sodium chloride - g / dm^3")
p5 <- qplot(data =pf, x = free.sulfur.dioxide) + xlab("free sulfur dioxide (mg / dm^3)")
p6 <- qplot(data =pf, x = total.sulfur.dioxide) + xlab("total sulfur dioxide (mg / dm^3)")
p7 <- qplot(data =pf, x = density) + xlab("density (g / cm^3)")
p8 <- qplot(data =pf, x = pH) + xlab("pH")
p9 <- qplot(data =pf, x = sulphates) +xlab("sulphates (potassium sulphate - g / dm3)")
p10 <- qplot(data =pf, x = alcohol) + xlab("alcohol (% by volume)")
p11 <- qplot(data =pf, x = quality)+ geom_bar()  + xlab("quality (score between 0 and 10)")
p12 <- qplot(data =pf, x = quality_cat) 
```

The first distribution we are going to analyse is about volatile acidity, citric.acid and residual of sugar. 

```{r}
grid.arrange(p1,p2,p3, ncol=1)
```

For the first two we can notice that some values are present in the tail. 
This is why we have added a boxplot to better investigate outliers both for 
acidity and citric acid. Boxplot can be very useful to visualize 
information that can be more difficult to find in a histogram. As you can 
notice, with this other plot type is much easier to have an idea of how many 
outliers we have for each tail.

```{r}
p1a <- ggplot(pf, aes(y = volatile.acidity, x = 1)) +
        geom_boxplot(outlier.color = "red") + coord_flip() +
  xlab("acetic acid - g / dm^3")

grid.arrange(p1, p1a, ncol=1)

```


```{r}
p21 <- ggplot(pf, aes(y = citric.acid, x = 1)) + 
  geom_boxplot(outlier.color = "red") + coord_flip()+
 xlab("citric acid (g / dm^3)")

grid.arrange(p2, p21, ncol=1)
```

Back to the distribution analysis: here for chlorides, free and total sulfur. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(p4,p5,p6, ncol= 1)
```

A closer analysis to the outliers in chlorides and free sulfur dioxide:

```{r}
p41 <- ggplot(pf, aes(y = chlorides, x = 1)) +
        geom_boxplot(outlier.color = "red") + coord_flip()+
   xlab("chlorides (sodium chloride - g / dm^3")

grid.arrange(p4, p41, ncol=1)

```

```{r}
p51 <- ggplot(pf, aes(y = free.sulfur.dioxide, x = 1)) +
        geom_boxplot(outlier.color = "red") + coord_flip()+
  xlab("free sulfur dioxide (mg / dm^3)")

grid.arrange(p5, p51, ncol=1)
```

Next chunk of distribution is about density, pH and sulphates. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(p7,p8,p9, ncol= 1)
```

Let's have a better look at the outliers for pH and sulphates.

```{r}
p81 <- ggplot(pf, aes(y = pH, x = 1)) +
        geom_boxplot(outlier.color = "red") + coord_flip()

grid.arrange(p8, p81, ncol=1)
```

```{r}
p91 <- ggplot(pf, aes(y = sulphates, x = 1)) +
        geom_boxplot(outlier.color = "red") + coord_flip() +
  xlab("sulphates (potassium sulphate - g / dm3)")

grid.arrange(p9, p91, ncol=1)
```

Then, the last distributions about alcohol, quality and the new variable
quality_cat. Here no outliers are identified. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(p10,p11,p12, ncol = 1)
```

# Univariate Analysis

### What is the structure of your dataset?
The dataset contains 11 features, all numerical, and one variable called quality, 
that represents how good or bad each wine is. A new categorical variable, 
quality_cat, has been added to the dataset to differentiate between Low, Medium 
and High quality instead of considering a full scale from 1 to 9.

### What is/are the main feature(s) of interest in your dataset?
From a first analysis we can notice that several variables seem of interest: 
alcohol level, residual of sugar and volatile acidity show some interesting 
pattern.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
In this first part we have just analysed single variables. In the bivariate 
analysis we might find interesting patterns between several variables that were 
not clearly identified here.

### Did you create any new variables from existing variables in the dataset?
Yes, quality_cat is a new variable derived from quality. It has value:

- "Low" if quality is less or equal to 4,

- "Medium" if quality is between 5 and 7 included,

- "High" if quality is higher or equal to 8.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
Yes, from what we have seen from histogram several distributions have outliers:
for all of those we have decided to plot again their histogram next to a boxplot.
This way identifying outliers is much easier than with a simple histogram.

# Bivariate Plots Section

The below plot represents the distribution of alcohol level for High, Low and 
Medium quality. As we can see, the majority of wines have a middle quality, 
while few are very good or bad.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(data =pf, x = alcohol)+ facet_wrap(~quality_cat) +xlab("alcohol (% by volume)")
```

In the below plot we show graphically how many values each category contains.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(data =pf, x = quality)+facet_wrap(~quality_cat)
```


To have a better idea about the differences between quality categories, we have 
done some boxplot. The idea is to find out some discrepancies in mean and median
between categories for each variables. Let's see if this is the case. 

The first boxplot below shows the alcohol distribution: as we can easily notice 
there is quite a difference between the average and median alcool level in 
High and Poor/Medium wine.

```{r}
qplot(y = alcohol,x= quality_cat, data = pf,
      geom = 'boxplot')+
    stat_summary(fun.y=mean, geom="point", shape=23, size=4)+
  ylab("alcohol (% by volume)")
```

Volatile acidity boxplot: here the difference between High and Low is less clear
for the median, but we can notice that Low wine have a higher mean compared to 
High ones. At the same time we can see that the distribution for Low is higher 
than High wine. Several outliers are present for Medium wine. 

```{r}
qplot(y = volatile.acidity,x= quality_cat, data = pf,
      geom = 'boxplot')+  coord_cartesian(ylim = c(0, 0.8))+
  stat_summary(fun.y=mean, geom="point", shape=23, size=4)+ 
  ylab("volatile acidity (acetic acid - g / dm^3)")
```

For fixed acidity and density, it is more complex to identify a clear difference
in mean and median in each quality.

```{r}
qplot(y = fixed.acidity,x= quality_cat, data = pf,
      geom = 'boxplot') +coord_cartesian(ylim = c(3, 10))+ 
  stat_summary(fun.y=mean, geom="point", shape=23, size=4)+ 
  ylab("fixed acidity (tartaric acid - g / dm^3)")
```

```{r}
qplot(y = density,x= quality_cat, data = pf,
      geom = 'boxplot') + coord_cartesian(ylim = c(0.98, 1.01))+ 
  stat_summary(fun.y=mean, geom="point", shape=23, size=4)+
  ylab("density (g / cm^3)")
```

The last variable that we are analysing is residual.sugar. Here we can denote differences in the distribution and median values for all categories.

```{r}
qplot(y = residual.sugar,x= quality_cat, data = pf,
      geom = 'boxplot') + coord_cartesian(ylim = c(0, 25))+  
  stat_summary(fun.y=mean, geom="point", shape=23, size=4)+
  ylab("residual sugar (g / dm^3)")
```


First of all let's create a plot that contains all combination of variables with
their plot and their correlation. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
opts_chunk$set(comment=NA, fig.width=10, fig.height=10)

```

```{r}
ggpairs(pf)
```

```{r}
opts_chunk$set(comment=NA, fig.width=6, fig.height=6)
```

From the plot above we have decided to focus on the following combiantion of variables:

- Density and residual sugar corr 0.839

- density and alcohol corr -0.78

- total.sulfur.dioxide and free.sulfur.dioxide corr 0.616

- density and total.sulfur.dioxide corr 0.53

- alcohol and residual sugar corr -0.451

- ph and fixed.acidity corr -0.42

- total.sulfur.dioxide and residual sugar corr 0.401

Below we will analyse the ones with the higher correlation. 

Let's start by density and residual sugar: 
in the plot below the trend is identified by the blue line.

```{r}
qplot(y = density, x =  residual.sugar,  data = pf)+ geom_point() + 
  xlim(c(0,20))+
  geom_point(alpha = 1/5, position  = position_jitter(), color = "orange") + 
  geom_smooth() +xlab("residual sugar (g / dm^3)")+
  ylab("density (g / cm^3)")
```

As well as for the plot of alcohol vs density we can see a linear trend - 
understanding why they are negative correlated. No particular transformation 
have been applied, just some outliers were removed.

```{r}
qplot(y = density, x = alcohol, data = pf)+ ylim(c(0.98,1.01))+
  geom_point(alpha = 1/5, position  = position_jitter(), color = "orange") + 
  geom_smooth() +xlab("alcohol (% by volume)") + ylab("density (g / cm^3)")

```

For total vs free sulfur we can also see that there is a relationship. Here on 
the x asis a square root transformation has been applied as well as some outliers
have been removed from the visualization. 


```{r}
qplot(y = total.sulfur.dioxide, x =  free.sulfur.dioxide, data = pf)+ 
  xlim(c(0,200))+
  geom_point(alpha = 1/5, position  = position_jitter(), color = "orange") + 
  coord_trans(x = "sqrt") + geom_smooth()+
  xlab("free sulfur dioxide (mg / dm^3)") + 
  ylab("total sulfur dioxide (mg / dm^3)")
```

Also in this plot we can clearly notice a linear trend, confirming what we saw 
in the initial plot and correlation value. 

```{r}
qplot(y = density, x =total.sulfur.dioxide , data = pf) + xlim(c(0,300))+ 
  ylim(c(0.95,1.05))+ 
  geom_point(alpha = 1/5, position  = position_jitter(), color = "orange") + 
  geom_smooth() + xlab("total sulfur dioxide (mg / dm^3)") +
  ylab("density (g / cm^3)")

```

In the next step of our analysis we are going to consider some variables together. 
We are starting with alcool and residual sugar: we are grouping  values for each
unit of alcohool (8,9, 10 etc) and then calculate the mean and median for those 
values.

```{r}
pf.wine_by_alcohol<- pf %>%
  group_by(round(alcohol))  %>%
  summarise(sugar_mean = mean(residual.sugar),
                         sugar_median = median(residual.sugar),
            n= n())%>%
  ungroup() 

tail(pf.wine_by_alcohol)
```

The below plot is about the mean sugar residual for alcohol levels grouped 
together. This clearly shows a peak of sugar mean for 9 degree of alcohol, which 
then decreases as the alcohol increases.

```{r}
ggplot(aes(x= `round(alcohol)`, y= sugar_mean), data= pf.wine_by_alcohol) +
  geom_line()

```

Here below the scatterplot for sugar vs alcohol and their trend. 

```{r}
ggplot(aes(y= residual.sugar, x= alcohol), data= pf) + 
  geom_point(alpha = 1/5, position  = position_jitter(), color = "orange")+
  ylim(0,30)+
  geom_smooth() +ylab("residual sugar (g / dm^3)") +xlab("alcohol (% by volume)")

```

Another analysis we are doing is grouping together density to calculate the mean 
and median residual sugar. 

```{r}
pf.wine_by_density<- pf %>%
  #filter(subset(pf, pf$quality_cat!='Medium')) %>%
  group_by(round(density, digit=3))  %>%
  summarise(sugar_mean = mean(residual.sugar),
                         sugar_median = median(residual.sugar),
            n= n())%>%
  ungroup() 

tail(pf.wine_by_density)

```

This plot represents density vs sugar mean value. A clear linear trend is 
identified here, confirming what we saw in the scatterplot above. 

```{r}
ggplot(aes(x= `round(density, digit = 3)`, y= sugar_mean), 
       data= pf.wine_by_density) +
  geom_line()

```

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
First of all we have started our bivariate analysis calculating the distribution, 
plot and correlations for all combination of variables. Considering that we did 
not know much of this dataset in advance, this gave us the opportunity to focus 
our initial analysis on some of the features. In particular, we have analysed:

- Density and residual sugar 

- density and alcohol corr 

- total.sulfur.dioxide and free.sulfur.dioxide corr 

- density and total.sulfur.dioxide corr.

For all of those we have found a linear trend that was easy to identify after 
having done some little adaptation to the plot. 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
We have also analysed alcohol and sugar values: to do so we have decided to group 
alcohol levels by unit and calculate for each mean and median. The plot clearly 
shows that there is a relationship between those two variables. Higher the level 
of alcohol, lower the sugar level, with a peak around 9. This confirmed what we 
knew about alcohol and sugar level. 

### What was the strongest relationship you found?
Between density and sugar level we have found a strong correlation of 0.8, 
confirmed both by the scatterplot and by the group analysis. 

# Multivariate Plots Section

In this analysis we will take the main variables we considered before and plot 
them against their quality category. Here we are particularly interested to see 
if there is a clear division between High and Low wine in each plot. In all plot 
below we have omitted all medium categories values, showing in yellow the points
about low quality and in orange the high ones.  

First let's look at density vs residual sugar. Clearly we can see that we could 
separate almost linearly those two categories: 

- very good wine appear to have mostly low level of residual of sugar and low 
density

- not good wine have instead a higher level of density even for higher level of 
sugar.
```{r echo=FALSE, Multivariate_Plots}
qplot(y = density, x= residual.sugar, data = subset(pf, pf$quality_cat!='Medium'), 
      color= quality_cat)+
geom_point(alpha = 0.15, size = 2, position = 'jitter') + 
  scale_color_brewer(type = 'div', palette = 'RdYlBu', 
                     guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  xlab("residual sugar (g / dm^3)") + ylab("density (g / cm^3)")
```

In the next plot we are focusing on density vs alcohol levels. As before, 
we could clearly separate the values between High and Low quality of wine. Here 
a higher quality means higher level of alcool and lower density.

```{r echo=FALSE}
qplot(y = density, x= alcohol, data = subset(pf, pf$quality_cat!='Medium'), 
      color= quality_cat)+
geom_point(alpha = 0.15, size = 2, position = 'jitter') + 
  scale_color_brewer(type = 'div', palette = 'RdYlBu', 
                     guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2)))+
  ylab("density (g / cm^3)") + xlab("alcohol (% by volume)")

```

In the below plot we have analysed total vs free sulfur dioxide. It's again 
possible to separate - not linearly this time - higher vs lower quality. 


```{r echo=FALSE}
qplot(y = total.sulfur.dioxide, x= free.sulfur.dioxide, 
      data = subset(pf, pf$quality_cat!='Medium'), color= quality_cat)+ 
  xlim(c(0,100))+ geom_point(alpha = 0.15, size = 2, position = 'jitter') + 
  scale_color_brewer(type = 'div', palette = 'RdYlBu', 
                     guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2)))+
  xlab("free sulfur dioxide (mg / dm^3)") + 
  ylab("total sulfur dioxide (mg / dm^3)")

```

Having a look at fixed.acidity vs pH it shows that we cannot always separate the
data between low and high. In the below plot, it is hard to define a rule, a lot
of points in the middle are about both low and high quality. 


```{r}
qplot(y = pH, x= fixed.acidity, data = subset(pf, pf$quality_cat!='Medium'), 
      color= quality_cat)+ 
geom_point(alpha = 0.15, size = 2, position = 'jitter') + 
  scale_color_brewer(type = 'div', palette = 'RdYlBu', 
                     guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2)))+
  xlab("fixed acidity (tartaric acid - g / dm^3)")

```

Another interesting plot is density vs total sulfur. Here we could separate with 
an hyperplane points that are about the two categories, having high quality wine 
usually with medium level of sulfur dioxide and not too high density.


```{r}
qplot(y = density, x= total.sulfur.dioxide, 
      data = subset(pf, pf$quality_cat!='Medium'), color= quality_cat)+ 
geom_point(alpha = 0.15, size = 2, position = 'jitter') + 
  scale_color_brewer(type = 'div', palette = 'RdYlBu', 
                     guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  xlab("density (g / cm^3)") + ylab("total sulfur dioxide (mg / dm^3)")

```

Let's now have a look at a similar analysis we did already before. 
Here we will focus on density and residual sugar mean considering as well 
different quality. 


```{r}
pf.fc_by_density_quality<- pf %>%
  group_by(round(density, digit=3), quality_cat)  %>%
  summarise(sugar_mean = mean(residual.sugar),
                         sugar_median = median(residual.sugar),
            n= n())%>%
  ungroup() %>%
  arrange(quality_cat)

head(pf.fc_by_density_quality)

```

As we can see from the plot below, the mean value of sugar are different between
low and high quality. 

```{r}
ggplot(aes(x= `round(density, digit = 3)`, y= sugar_mean), 
       data= subset(pf.fc_by_density_quality, quality_cat!='Medium')) +
geom_line(aes(color= quality_cat))
```

At the same time we have grouped by alcohol, considering as well quality. 

```{r}
pf.wine_by_alcohol_quality<- pf %>%
  group_by(round(alcohol), quality_cat)  %>%
  summarise(sugar_mean = mean(residual.sugar),
                         sugar_median = median(residual.sugar),
            n= n())%>%
  ungroup()%>%
  arrange(quality_cat)

head(pf.wine_by_alcohol_quality)


```

Here as well, we can see a difference between the mean value of high and low 
qualities. 

```{r}
ggplot(aes(x= `round(alcohol)`, y= sugar_mean), 
       data= subset(pf.wine_by_alcohol_quality, quality_cat!='Medium')) +
geom_line(aes(color= quality_cat))
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
In this part of the analysis we focused mostly on analysing the relationship 
between Low and High quality wine. Considering that those two categories contain 
a similar amount of data, we have just taken  them in consideration, trying to 
find a clear pattern between the best and worst wine. It is clear that some 
variables play an important role in defining best and worst wine. Several plots,
like density vs alcohol or density vs residual sugar, show how we can clearly 
separate high vs low qualities. 

### Were there any interesting or surprising interactions between features?
It was interesting to see that some variables separate better than others the two
categories. At the same time, it was interesting to confirm some trends that we 
noticed already in the bivariate analysis. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
qplot(y = alcohol,x= quality_cat, data = pf,
      geom = 'boxplot', main = "Boxplot alcohol by quality cat")+
    stat_summary(fun.y=mean, geom="point", shape=23, size=4)+
  xlab("quality category")+
  ylab("alcohol (% by volume)")
```

```{r}
summary(pf$alcohol)
```

### Description One
This plot is a boxplot of alcohol for each different quality low, medium, high. 
This plot is very informative because we can see the distribution for each class, 
its mean and median and if outliers are present. It also shows us a clear 
difference between high level wine and low/medium ones.

### Plot Two
```{r echo=FALSE, Plot_Two}
qplot(y = density, x= residual.sugar, data = subset(pf, pf$quality_cat!='Medium'), 
      main = "Density vs residual sugar in low and high wine types", 
      color= quality_cat)+ 
  geom_point(alpha = 0.15, size = 2, position = 'jitter') + 
  scale_color_brewer(type = 'div', palette = 'RdYlBu', 
                     guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2)))+
  xlab("residual sugar (g / dm^3)") + 
  ylab("density (g / cm^3)")
```

### Description Two
This scatterplot gives us information about the relationship of density vs
residual sugar. It is very informative because we can clearly divide the two 
classes of wine Low and High with a simple line. It gives us an insight on the 
possible use of a classifier. A SVM could be used to find a hyperplane to divide 
the two classes of data. 

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x= `round(alcohol)`, y= sugar_mean), 
       data= subset(pf.wine_by_alcohol_quality, quality_cat!='Medium')) + 
  ggtitle("Average residual sugar per alcohol in low and high wine types")+
geom_line(aes(color= quality_cat))
```

### Description Three

The third graph that I am using here it is a line plot about average residual of 
sugar for alcohol unit. Here we are plotting just information about high and low
quality and we can see a difference till degree 13 for those two types of wine.

------

# Reflection

It was very interesting analysing this dataset. I like drinking wine when I hang
out with friends and I could tell when a wine is good or not. Till now I was not 
really aware of all chemical properties that we can find in wine and how they do 
play a role in its quality. 
In the beginning it was complicated to start, since I didn't know much about 
this dataset . Once I got more familiar with it, I could find more interesting 
things.
Now I have a better idea on what influences  having a good wine instead of a bad 
one. In chemical terms, I can definitely find a correlation between density 
vs alcohol, residual sugar and total.sulfur.dioxide. At the same time the 
relationship between alcohol and residual sugar is clear and shows a clear pattern.

For future work, it would be interesting to explore more the analysis on chemical 
properties, but we could add more value having more generic information about the data
, like type of wine, where it grows, temperature, etc. Like that the analysis
would be more complete, not just focusing on chemical compounds that not everyone
is well aware of, but as well more tangible facts. 